---
title: Data Analysis

author: "Tung Tran"
date: "2025-03-16"
output: html_document
---

Question : My research question is to see if the predictors can adequately predict the sufficiency of removing invasive clams off boats.


       J. Filename: hot_water_removal.csv      
     
      Contents: Includes information about the max temperature and duration of water sprayed on each boat section by each decontaminator participant. Also includes the location where the decontamination took place, duration of the decontamination, steps that were taken, decontaminaiton unit type, and sufficiency rankings.
	
       
       Name: prep_duration
       Description: duration (HH:MM:SS) a decontaminator spent preping for their water decontamination process (the time in between hand inspection and water deocntamination)
       
    J. Name: decon_duration
       Description:duration (HH:MM:SS) of a decontaminators water decontamination step 
       
       . Name: inspection_duration
       Description: duration (HH:MM:SS) of a participants hand inspection step
       
       
```{r}


hot_water_removal = read.csv("C:/Users/avist/Desktop/raw data/hot_water_removal.csv")

#Theres some NA values so im cleaning it
hot_water_removal_clean = na.omit(hot_water_removal)

#1 for sufficient, 0 for not
hot_water_removal_clean$overall_sufficiency = ifelse(hot_water_removal_clean$overall_sufficiency == "Sufficient", 1, 0)
hot_water_removal_clean$overall_sufficiency


```



```{r}

#The code is transforming time-based variables (in the format HH:MM:SS) into numeric values representing the total number of seconds for each corresponding variable. This makes it easier to analyze and model these duration in subsequent statistical analysis.


hot_water_removal_clean$prep_duration_seconds = as.numeric(as.difftime(hot_water_removal_clean$prep_duration, format = "%H:%M:%S"))
hot_water_removal_clean$decon_duration_seconds = as.numeric(as.difftime(hot_water_removal_clean$decon_duration, format = "%H:%M:%S"))
hot_water_removal_clean$inspection_duration_seconds = as.numeric(as.difftime(hot_water_removal_clean$inspection_duration, format = "%H:%M:%S"))


#Seeing which variables have correlation ( -1 to 1 ) 
cor_matrix = cor(hot_water_removal_clean[, c("prep_duration_seconds", "decon_duration_seconds", "inspection_duration_seconds", "max_temp")])

#It seems like only MAX temp has a low correlation to the other variables


boxplot(max_temp ~ overall_sufficiency, data = hot_water_removal_clean )
#Comparing the median, the suffiencent seems to have a higher median


hist(hot_water_removal_clean$max_temp, breaks = 20)
#this is one of the challenge since its unusual because the tempt of 0. Large sum of 0 is an indication (our data is inaccurate because having 0 as a max temperature does not make sense)
mean(hot_water_removal_clean$max_temp == 0)

```


High correlations:

total_duration_seconds and prep_duration_seconds have a strong positive correlation of 0.6338. This suggests that as one of these variables increases, the other tends to increase as well.
total_duration_seconds and decon_duration_seconds have a moderate positive correlation of 0.7181, meaning they are somewhat related.
decon_duration_seconds and inspection_duration_seconds show a moderate positive correlation of 0.4341.


Low or weak correlations:

The correlations between max_temp and other variables (such as total_duration_seconds, prep_duration_seconds, and decon_duration_seconds) are low. For example, the correlation between total_duration_seconds and max_temp is only 0.1979, suggesting a weak positive relationship between them.


Near-zero correlations:

prep_duration_seconds and decon_duration_seconds have a very weak negative correlation of -0.0114, indicating almost no relationship between these two variables.



#test and training 
```{r}
library(glmnet)
library(caret)

# Split data
train_index = sample(1:nrow(hot_water_removal_clean), 0.8 * nrow(hot_water_removal_clean))
train_data = hot_water_removal_clean[train_index, ]
test_data = hot_water_removal_clean[-train_index, ]

# Create design matrices
train_x = model.matrix(overall_sufficiency ~  prep_duration_seconds + 
                       decon_duration_seconds + inspection_duration_seconds + max_temp, data = train_data)

test_x = model.matrix(overall_sufficiency ~  prep_duration_seconds + 
                      decon_duration_seconds + inspection_duration_seconds + max_temp, data = test_data)

# Stepwise logistic regression
model = glm(overall_sufficiency ~ prep_duration_seconds + 
            decon_duration_seconds + inspection_duration_seconds + max_temp, 
            data = train_data, family = "binomial")
model

step_model = step(model, direction = "both")
coef(step_model)
# Lasso with cross-validation
cv_lasso = cv.glmnet(train_x, train_data$overall_sufficiency, alpha = 1, family = "binomial")
best_lambda_lasso = cv_lasso$lambda.min
lasso_model = glmnet(train_x, train_data$overall_sufficiency, alpha = 1, lambda = best_lambda_lasso, family = "binomial")
coef(lasso_model)
# Ridge with cross-validation
cv_ridge = cv.glmnet(train_x, train_data$overall_sufficiency, alpha = 0, family = "binomial")
best_lambda_ridge = cv_ridge$lambda.min
ridge_model = glmnet(train_x, train_data$overall_sufficiency, alpha = 0, lambda = best_lambda_ridge, family = "binomial")
coef(ridge_model)
# Print optimal lambdas
cat("Best lambda (Lasso):", best_lambda_lasso, "\n")
cat("Best lambda (Ridge):", best_lambda_ridge, "\n")



# Make predictions
predictions_ridge_test = predict(ridge_model, newx = test_x, type = "response", s = best_lambda_ridge)
predictions_lasso_test = predict(lasso_model, newx = test_x, type = "response", s = best_lambda_lasso)
predictions_step_test = predict(step_model, newdata = test_data, type = "response")


# Thresholding
threshold = 0.5
predictions_ridge_binary = ifelse(predictions_ridge_test > threshold, 1, 0)
predictions_lasso_binary = ifelse(predictions_lasso_test > threshold, 1, 0)
predictions_step_binary = ifelse(predictions_step_test > threshold, 1, 0)



# Confusion matrices
confusion_ridge = confusionMatrix(as.factor(predictions_ridge_binary), as.factor(actual_labels))
confusion_lasso = confusionMatrix(as.factor(predictions_lasso_binary), as.factor(actual_labels))
confusion_step = confusionMatrix(as.factor(predictions_step_binary), as.factor(actual_labels))

# Output
confusion_ridge
confusion_lasso
confusion_step



```


```{r}


all_predictors = c("prep_duration_seconds", 
                   "decon_duration_seconds", "inspection_duration_seconds", "max_temp")

## ---- STEPWISE ----
cat("\n--- Stepwise: Dropped Predictors ---\n")
stepwise_coefs = coef(step_model)
stepwise_dropped = setdiff(all_predictors, names(stepwise_coefs)[stepwise_coefs != 0])
cat("Dropped by Stepwise:\n")
print(stepwise_dropped)


## ---- LASSO ----
cat("\n--- Lasso: Shrunk Predictors (coef = 0) ---\n")
lasso_coefs = coef(cv_lasso, s = "lambda.min")
lasso_coefs_matrix = as.matrix(lasso_coefs)  # Convert to regular matrix
lasso_dropped = rownames(lasso_coefs_matrix)[lasso_coefs_matrix == 0]
lasso_dropped = setdiff(lasso_dropped, "(Intercept)")  # remove intercept
cat("Shrunk by Lasso:\n")
print(lasso_dropped)


#---- RIDGE ----
cat("\n--- Ridge: Shrunk Predictors (|coef| < 0.01) ---\n")
ridge_coefs = coef(cv_ridge, s = "lambda.min")
ridge_coefs_matrix = as.matrix(ridge_coefs)  # Convert to regular matrix
ridge_values = ridge_coefs_matrix[, 1]  # Extract coefficients
ridge_names = rownames(ridge_coefs_matrix)

# Define a threshold for "shrunk" coefficients (e.g., smaller than 0.01)
threshold = 0.01
shrunk_idx = which(abs(ridge_values) < threshold & ridge_names != "(Intercept)")
ridge_shrunk = ridge_names[shrunk_idx]
cat("Shrunk by Ridge (|coef| < 0.01):\n")
print(ridge_shrunk)



```


```{r}


plot(cv_lasso$glmnet.fit, xvar = "lambda", label = TRUE)
title("Lasso Coefficient Path")

plot(cv_ridge$glmnet.fit, xvar = "lambda", label = TRUE)
title("Ridge Coefficient Path")

coef(ridge_model)
```
